>

### 2019-04-11 paper/spinifex_paper.Rmd
Writing changes after meeting with Julie.

  A recent paper [@wang_mapping_2018] visualizes the sensitivity of the hadronic experiments to nucleon structure. Sensitivity was characterized in non-linear 3D embeddings of the first 10 principal components. This research applies manual tours to this data showing that manual tours resolve information about cluster structure and identify outliers, information that is orthogonal to the original static viewing plane.
keywords: manual tour, guided tour, grand tour, projection pursuit, high dimensional data, multivariate data, data visualization, statistical graphics, data science
preamble: >
  % preamble packages
link-citations: true
fig_caption: yes
toc: false
bibliography: RJrefs.bib
output: rticles::rjournal_article
  #bookdown::pdf_book:
   # base_format: rticles::rjournal_article #see sandbox for hard code solution.
---
\bibliography{RJrefs}
<!-- note that Di/Ursula have different setup and options. use static images.  -->
```{r setup_paper, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center", echo = TRUE, collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  cache = F, cache.lazy = FALSE
)
library(spinifex)
library(ggplot2)
```

@@

<!-- Tours -->
A tour is a multivariate data analysis technique in which a sequence of linear (orthogonal) projections are viewed as an animation while the orientation of the projection basis is rotated across time. Each frame of the sequence corresponds to a small change in the projection for a smooth transition that perseveres continuity. This is a crucial tool extending the dimensionality of visualization which is a large boon as datasets and models exist in increasing cardinality. The package \CRANpkg{tourr} [@wickham_tourr_2011] builds a platform for generating tour paths and applying to a wide number of display geometrics in **base** graphics.

<!-- Manual tour -->
While there are numerous methods that generate tour paths, this research focuses on the manual tour. The manual tour was described in @cook_manual_1997 and allows a user to control the projection coefficients of a selected variable in a 2D projection. The manipulation of these coefficients allows the analyst to explore how sensitive the projections structure is to these changes. This makes manual tours particularly useful once a feature of interest has been identified, for example, with the use of a guided tour [@hurley_analyzing_1990]. Rather than exploring the large phase space of the data, a guided tour selects a very specific path via projection pursuit [@diaconis_asymptotics_1984], the optimization of an index function on the projection via a hill climbing algorithm. This allows guided tours to rapidly identify interesting projection features given the relatively large parameter-space. After a projection of interest is identified an analyst can then use the "finer brush" of the manual tour, by controlling the contributions of individual variables.

<!-- Display -->
\CRANpkg{spinifex} utilizes two new animation packages, \CRANpkg{plotly} [@sievert_plotly_2018] and \CRANpkg{gganimate} [@pedersen_gganimate:_2019], to display tours, manual or other saved tours. From a given projection, the user can choose which variable to control, and the animation sequence is generated to remove the variable from the projection, and then extend its contribution to be the sole variable in one direction. This allows the viewer to assess the change in structure induced in the projection by the variable's contribution.

<!-- Paper outline -->
The paper is organized as follows. Section \ref{sec:algorithm} explains the algorithm using a toy dataset. Section \ref{sec:display} discussed the display of the animation after the path has been generated. Section \ref{sec:application} illustrates how this can be used for sensitivity analysis applied to contemporary high energy physics. The last section, \ref{sec:discussion} summarizes the work and discusses future research.

@@

<!-- describe data. -->
The flea data, originally from @lubischew_use_1962, made available in **tourr**  is used to illustrate the algorithm. The data contains 74 observations across 6 variables, physical measurements of the flea beetles. Each observation belonging to one of three species.

<!-- Projection basis -->
The data is defined. A basis set (ideally orienting to an interesting feature) must be provided as an initial orientation. One way to identify a projection containing interesting features is to apply a guided tour [@cook_interactive_2007]. In a guided tour the projection sequence is selected by optimizing an index function via hill-climbing on the projection space. In this case, the holes index is selected and applied to standardized flea data. The holes index is maximized when the projected observations are furthest from the center. Figure \ref{fig:step0} shows a locally optimized projection of the data. The left panel displays the reference axes of the projection basis, a visual indication of the magnitude and direction each variable contributed to the projection. The right panel shows the data as projected through the basis set described by the reference axes (left). Data points are colored and given shape according to the species (while the guided tour was unsupervised with this information).

```{r step0, echo=F, results='hide', out.width='70%',                             fig.cap = "Basis reference axes (left) and projected data (right) of standardized flea data. Data points color and shape are mapped to beetle species. Basis identified by a holes-index guided tour. The variables aede2 and tars1 contribute mostly orthogonal to the other variables."}
## Flea holes tour
f_dat  <- tourr::rescale(flea[,1:6])
f_cat  <- factor(flea$species)
f_path <- save_history(f_dat, guided_tour(holes()))
f_bas  <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
f_mvar <- 5
f_msp  <- create_manip_space(basis = f_bas, manip_var = f_mvar)
f_proj <- data.frame(tourr::rescale(f_dat %*% f_msp[, 1:2]))

@@

(ref:JetClusterBad-cap) A radial manual tour of PC4 within the jet cluster. Colored by experiment type: ATLAS7new in green and ATLAS7old in orange. Data from ATLAS7new (green) spans mostly the same space as ALTLAS7old (orange) with no evident difference in cluster structure across varying contributions of PC3. A dynamic version can be viewed at https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc3/.

```{r JetClusterBad, echo=F, warning=F, fig.height=7.2, fig.width=6, out.height='7.2in', out.width='6in', fig.cap = "(ref:JetClusterBad-cap)"}
# Jet cluster, Fig 7 of cook_dynalical_2018, subset to ATLAS7old and ATLAS7new
jet_mvar  <- 3
jet_mtour <- manual_tour(basis = jet_bas, manip_var = jet_mvar, angle = jet_ang)

@@

<!-- introduce DIS cluster -->
A different space is used to explore the DIS cluster; specifically the first six principal components, which explain 48% of the variation contained within the aggregated data[@cook_dynamical_2018]. Radial manual tours are performed on PC6 and PC2 in figure \ref{fig:DISclusterGood} and figure \ref{fig:DISclusterBad} respectively.

<!-- comparison of the DIS cluster -->
The selection of the manip variable is important as the manipulation spaces convey substantially different information. The manual tour of PC6 offers information about the dimensionality, shape, and orientations of the different experiment classes. whereas manipulating the contributions of PC2 only shows a subset of the dimensionality and shape information. Manipulating the contributions of PC6 turned out to be much more insightful than that of PC2. This result might seem counter-intuitive at first as PC2 should explain much more of the variation in the data. However, features and structure in the data regularly reside in finer detail that can be lost when looking only at static projections.

@@

<!-- summary and segue  -->
Tours, the dynamic linear projection of multivariate data, play an important role in data visualization; they extending the dimensionality of visuals while data- and parameter-spaces become ever larger. This research has modified the algorithm producing manual tours and has made this functionality available in package **spinifex**. The package adds to **tourr**, extending the graphics offerings that can be used to display tours.

Radial manual tours were applied to a dataset across different experiments of hadronic collisions. The importance of selecting the correct variable to manipulate is demonstrated by comparing tours of varying quality. The manual tours convey a better picture of the structure of the clustering than static linear or non-linear projections are able to. Giving the full contribution of the manipulation space to the manip var enables analysts to explore the sensitivity of the structure with the selected variable. This information can be used by physicists to which experiments are probing which parameter spaces, which can be indispensable for smarter planning of effort and funding.

@@

<!-- geometric and graphics display -->
The **tourr** package provides many other geometric displays with the `tourr::display_*()` family. These geometric options could be integrated into the **ggplot2** framework for display on **plotly** and **gganimate**. Additionally, the \CRANpkg{animation} package @xie_animation:_2018 could be implemented for another graphics framework. However, **animation** builds from **base** graphs while **spinifex** utilizes **ggplot2** graphics.

>>>>

  A recent paper [@wang_mapping_2018] visualized the sensitivity of the hadronic experiments to nucleon structure. Sensitivity was characterized in non-linear 3D embeddings of the first 10 principal components. This research applies manual tours to this data showing that manual tours resolve information about the structure of cluster distribution by bringing to light information that is orthogonal to static viewing planes.
keywords: manual tour, guided tour, grand tour, projection pursuit, high dimensional data, multivariate data, data visualization, statistical graphics, data science
preamble: >
  % preamble packages
link-citations: true
fig_caption: yes
toc: false
bibliography: RJrefs.bib
output: rticles::rjournal_article
  #bookdown::pdf_book:
   # base_format: rticles::rjournal_article #see sandbox for hard code solution.
---
\bibliography{RJrefs}
<!-- note that Di/Ursula have different setup and options and use static images.  -->
```{r setup_paper, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center", echo = TRUE, collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  cache = F, cache.lazy = FALSE
)
library(spinifex)
library(ggplot2)
```

@@

<!-- Tours -->
A tour is a multivariate data analysis technique in which a sequence of linear (orthogonal) projections are viewed as an animation while the orientation of the projection basis is rotated across time. Each frame of the sequence corresponds to a small change in the projection for a smooth transition that preserves continuity. Tours are a crucial tool extending the dimensionality of visualization which is increasingly important as datasets and models exist in increasing cardinality. The package \CRANpkg{tourr} [@wickham_tourr_2011] builds a platform for generating tour paths and applying to a wide number of display geometrics in **base** graphics.

<!-- Manual tour -->
While there are numerous methods that generate tour paths, this research focuses on the manual tour. The manual tour was described in @cook_manual_1997 and allows a user to control the projection coefficients of a selected variable in a 2D projection. The manipulation of these coefficients allows the analyst to explore how sensitive the projection structure is to these manipulations. As manual tours operate on only one variable at a time, they particularly useful once a feature of interest has been identified. One way to identify "interesting" features is with the use of a guided tour [@hurley_analyzing_1990]. Guided tours select a very specific path, that which approachs a projection that optimizes an objective-function, similar to simulated annealing [@kirkpatrick_optimization_1983]. The direct optimization of a function allows guided tours to rapidly identify interesting projection features given the relatively large parameter-space. After a projection of interest is identified an analyst can then use the "finer brush" of the manual tour, by controlling the contributions of individual variables to explore the sensitivity they have to the projection structure.

<!-- Contributions -->
This paper offers an implementation of manual tours with the package
\CRANpkg{spinifex}. This package also exends the graphics display of **tourr** to the animation frameworks offered by \CRANpkg{plotly} [@sievert_plotly_2018] and \CRANpkg{gganimate} [@pedersen_gganimate:_2019]. The algorithm used to perform a radial manual tour is demonstrated on a toy data set. Followed by an application and disscussion of the radial manual tour on an aggregation of high energy physics experiments in a large multivariate space.

<!-- Paper outline -->
The paper is organized as follows. Section \ref{sec:algorithm} explains the algorithm using a toy dataset. Section \ref{sec:display} discusses the display of the animation after the path has been generated. Section \ref{sec:application} illustrates how this can be used for sensitivity analysis applied to contemporary high energy physics. The last section, \ref{sec:discussion} summarizes the work and discusses future research.

@@

<!-- describe data. -->
The flea data, originally from @lubischew_use_1962, made available in **tourr**  is used to illustrate the algorithm. The data contains 74 observations across 6 variables, which are physical measurements of the flea beetles. Each observation belong to one of three species.

<!-- Projection basis -->
The data is defined. A basis set (ideally orienting to an interesting feature) must be provided as an initial orientation. One way to identify a projection containing interesting features is to apply a guided tour. In a guided tour, the projection sequence is selected by optimizing an index function via hill-climbing on the projection space. In this case, the holes index is selected and applied to standardized flea data. The holes index is maximized when the projected observations are furthest from the center. Figure \ref{fig:step0} shows a locally optimized projection of the data. The left panel displays the reference axes of the projection basis, a visual indication of the magnitude and direction each variable contributed to the projection. The right panel shows the data as projected through the basis set described by the reference axes (left). Data points are colored and given shape according to the species (while the guided tour was unsupervised with this information).

```{r step0, echo=F, results='hide', out.width='70%',                             fig.cap = "Basis reference axes (left) and projected data (right) of standardized flea data. Data points color and shape are mapped to beetle species. The basis was identified by a holes-index guided tour. The variables aede2 and tars1 contribute mostly orthogonal to the other variables."}
## Flea holes tour
f_dat  <- tourr::rescale(flea[,1:6])
f_cat  <- factor(flea$species)
f_path <- save_history(f_dat, guided_tour(holes()))
f_bas  <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2)
f_mvar <- 5
f_msp  <- create_manip_space(basis = f_bas, manip_var = f_mvar)
f_proj <- data.frame(tourr::rescale(f_dat %*% f_msp[, 1:2]))

@@

```{r JetClusterBad, echo=F, warning=F, fig.height=7.2, fig.width=6, out.height='7.2in', out.width='6in',                                       fig.cap = "A radial manual tour of PC4 within the jet cluster. Colored by experiment type: ATLAS7new in green and ATLAS7old in orange. Data from ATLAS7new (green) spans mostly the same space as ALTLAS7old (orange) with no evident difference in cluster structure across varying contributions of PC3. A dynamic version can be viewed at https://nspyrison.netlify.com/thesis/jetcluster\\_manualtour\\_pc3/."}
# Jet cluster, Fig 7 of cook_dynalical_2018, subset to ATLAS7old and ATLAS7new
jet_mvar  <- 3
jet_mtour <- manual_tour(basis = jet_bas, manip_var = jet_mvar, angle = jet_ang)

@@

<!-- introduce DIS cluster -->
A different space is used to explore the DIS cluster; specifically the first six principal components, which explains 48% of the variation contained within the aggregated data [@cook_dynamical_2018]. Radial manual tours are performed on PC6 and PC2 in figure \ref{fig:DISclusterGood} and figure \ref{fig:DISclusterBad} respectively.

<!-- comparison of the DIS cluster -->
The selection of the manip variable is important as the manipulation spaces convey substantially different information. The manual tour of PC6 offers information about the dimensionality, shape, and orientations of the different experiment classes. Whereas manipulating the contributions of PC2 only shows a subset of the dimensionality and shape information. Manipulating the contributions of PC6 turned out to be much more insightful than that of PC2. This result might seem counter-intuitive at first as PC2 should explain much more of the variation in the data. However, features and structure in the data regularly reside in finer detail that can be lost when looking only at static projections.

@@

<!-- summary and segue  -->
Tours, which are dynamic linear projection of multivariate data, play an important role in data visualization; they extend the dimensionality of visuals while data- and parameter-spaces become ever larger. This research has modified the algorithm producing manual tours which and has made this functionality available in package **spinifex**. The package adds to **tourr**, extending the graphics offerings that can be used to display tours.

Radial manual tours were applied to a dataset across different experiments of hadronic collisions. The importance of selecting the correct variable to manipulate is demonstrated by comparing tours of varying quality. The manual tours convey a better picture of the structure of the clustering than static linear or non-linear projections. Giving the full contribution of the manipulation space to the manip var enables analysts to explore the sensitivity of the structure with the selected variable. This information can be used by physicists to identify which experiments are probing which parameter spaces, which can be indispensable for smarter planning of effort and funding.

@@

<!-- geometric and graphics display -->
The **tourr** package provides many other geometric displays with the `tourr::display_*()` family. These geometric options could be integrated into the **ggplot2** framework for display on **plotly** and **gganimate**. Additionally, the \CRANpkg{animation} package @xie_animation:_2018 could be implemented for another graphics framework. However, **animation** builds from **base** graphics while **spinifex** utilizes **ggplot2** graphics, a significant paradigm shift.
